---
title: "covariate_data"
output:
#  pdf_document:
#    toc: yes
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction
========================================================
Covariate analysis for 2015 rare data collected by Ohio State.

Distribution
========================================================
The git repo holding this code is available at:
https://github.com/puruckertom/rare_pollen

Computational environment
========================================================

```{r eval=FALSE, echo=FALSE}
#library(installr) #windows only currently
#updateR()
```
Version and installed libraries.
```{r eval=TRUE, echo=TRUE}
verbose_output = TRUE

if(verbose_output){
  print(Sys.info()[4])
  R.Version()$version.string
}
#check to see if needed packages are installed
list.of.packages <- c("ggplot2", "dplyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
#load relevant packages
library(MASS)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(reshape2)

if(verbose_output){
  print("list of loaded packages: ")
  print((.packages()))
}
```

```{r eval=TRUE, echo=FALSE}
if(Sys.info()[4]=="DZ2626UTPURUCKE"){
  rare_data <- "k:\\git\\rare_pollen\\"
} else if(Sys.info()[4]=="stp-air.local" || Sys.info()[4]=="stp-air"){
  rare_data <- path.expand("~/git/rare_pollen/")
} else if(Sys.info()[4]=="Erins-MacBook"){
  rare_data <- "~/Documents/GitHub/rare_pollen/"
} else
  rare_data <- paste(getwd(),"/",sep="")
  
osu_data <- paste(rare_data, "data_in/", sep="")
```

Analytical data
========================================================
Original file with observational data: 
https://github.com/puruckertom/rare_pollen/blob/master/data_in/rare_data_osu_2015.csv

```{r eval=TRUE, echo=TRUE}
##################
#the data sets
##################
#import raw data - everything
file.exists(paste(osu_data,"CDRC_2017EPA_FP_DBT.csv",sep=""))
osu_covariate <- read.table(paste(osu_data,"CDRC_2017EPA_FP_DBT.csv",sep=""), header = TRUE, sep = ",",strip.white=T,na.strings="")

```

####Cleaning the data

```{r eval=TRUE, echo=TRUE}
osu_covariate <- osu_covariate[,1:23]
colnames(osu_covariate)[c(4,6)] <- c("planting","per_corn")
osu_covariate$planting<- factor(osu_covariate$planting,labels=c("no","yes"))
neonics_cols <- which(grepl("DBT|FP",colnames(osu_covariate))& !grepl("sample_id",colnames(osu_covariate)))

#Create categorical vars which track whether any given neonic was detected
neonics_detected <- as.data.frame(lapply(osu_covariate[,neonics_cols],function(x) ifelse((grepl("<",x) | is.na(x)),0,1)))
colnames(neonics_detected) <- paste(colnames(neonics_detected),"detected",sep="_")
osu_covariate <- cbind(osu_covariate,neonics_detected)

#Create two single categorical vars which tracks whether ANY neonic was detected in dead bees or pollen
osu_covariate$DBT_any_detected <- ifelse(rowSums(neonics_detected[,grepl("DBT",colnames(neonics_detected))]) > 0,1,0)
osu_covariate$FP_any_detected <- ifelse(rowSums(neonics_detected[,grepl("FP",colnames(neonics_detected))]) > 0,1,0)


#convert neonic concs to numeric
osu_covariate[,neonics_cols] <- as.numeric(gsub("< ","",as.matrix(osu_covariate[,neonics_cols])))

#create variables for total neonic loads
osu_covariate$DBT_total_neonic <- rowSums(osu_covariate[,grepl("DBT",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

osu_covariate$FP_total_neonic <- rowSums(osu_covariate[,grepl("FP",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

neonics_cols <- c(neonics_cols,which(colnames(osu_covariate) %in% c("DBT_total_neonic","FP_total_neonic")))

#write.csv(osu_covariate,"data_in/CDRC_2017EPA_FP_DBT_clean.csv",row.names=FALSE) #output a csv for the cleaned data

#str(osu_covariate) #display summary of cleaned dataframe

```


Exploring the data
========================================================

<br>

#### Let's make a quick scan for correlations in the data....

<br><br>

##### What about correlations between mortality and any of the neonic concentrations?

<br><br>
```{r eval=TRUE, echo=TRUE}
for(i in 1:length(neonics_cols)){
  cor <- cor.test(osu_covariate$mortality, osu_covariate[,neonics_cols[i]])
  cat("\n",paste("Correlation between mortality and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))), ":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```
<br><br>

* So, we see strong positive correlations between mortality and two neonics, thiamethoxam and clothianidin, but for field pollen samples only. We may want to focus some analyses on these two chemicals. There is also a significant positive correlation for total pollen load of neonics
<br><br>

#####Now let's look at some correlations between the percent corn and the neonics
<br><br>

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(neonics_cols)){
  cor <- cor.test(osu_covariate$per_corn, osu_covariate[,neonics_cols[i]])
  cat("\n",paste("Correlation between % corn and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))),":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```
<br><br>

* It looks like there are near-significant postive correlations between % corn and these same two neonics (thiamethoxam and clothianidin) in pollen, as well as total pollen load. 
<br><br>

#####What if we just look at this relationship during defined corn planting periods?

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(neonics_cols)){
  cor <- cor.test(subset(osu_covariate,planting=="yes")$per_corn, subset(osu_covariate,planting=="yes")[,neonics_cols[i]])
  cat("\n",paste("Correlation between % corn and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))),":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```
<br>

* During planting periods, there are significant correlations between thiamethoxam and clothianidin in pollen, as well as total pollen load. 
<br><br>


Regression analysis - is mortality higher when neonic loads are higher?
==========================================
<br>

#####Lets start by looking at the just two neonics that we previously identified, as well as a total pollen neonic load:

```{r eval=TRUE, echo=TRUE}
interesting_neonics <- c("FP_thiamethoxam","FP_clothianidin","FP_total_neonic")

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste("mortality ~ ",interesting_neonics[i],
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",paste("Linear regression, ",interesting_neonics[i]," vs mortality:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  g <- ggplot(osu_covariate,aes_string(interesting_neonics[i],"mortality"))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x=paste(ifelse(substr(interesting_neonics[i],1,2)=="FP",
                        paste("field pollen",substr(interesting_neonics[i],4,20)),
                        paste("dead bees",substr(interesting_neonics[i],5,20))),"ng g-1"),
         y="Relative hive mortality")+geom_hline(yintercept=0)
  print(g)
}
```
<br><br>

Regression analysis - are there higher neonic loads when there's more corn nearby?
==========================================
<br><br>

#####Lets look at whether clothianidin, thiamethoxam, or total neonic load in pollen are linked to corn:
<br><br>

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste(interesting_neonics[i],"~ per_corn * planting",
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",paste("Linear regression, ",interesting_neonics[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
}

```
<br><br>

* Concentrations of these two neonics in pollen, as well as total neonic load, are increased when more corn is planted nearby, but ONLY during defined corn planting periods (see significant percent corn by planting interaction)

<br><br>

#####Let's graph how percent corn nearby affects field pollen concentrations of these neonics, by planting period
<br><br>

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(interesting_neonics)){
  g <- ggplot(osu_covariate,aes_string("per_corn",interesting_neonics[i]))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+facet_wrap(~planting,nrow=1,labeller=label_both)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) + coord_cartesian(ylim=c(0,150)) +
    labs(x="Percent corn within 2 km",y=paste(ifelse(substr(interesting_neonics[i],1,2)=="FP",
                        paste("field pollen",substr(interesting_neonics[i],4,20)),
                        paste("dead bees",substr(interesting_neonics[i],5,20))),"ng g-1"))
  print(g)
}

```

<br><br>

##### Let's do a check to see if there are similar relationships for any other neonics (in pollen or dead bees)
Note: output not shown <br>

```{r eval=FALSE, echo=TRUE}
neonics_to_eval <- colnames(osu_covariate)[which(grepl("FP|DBT",colnames(osu_covariate))& !grepl("FP_clothianidin|FP_thiamethoxam|FP_total_|detected|sample_id",colnames(osu_covariate)))]

for(i in 1:length(neonics_to_eval)){
  fm <- lm(as.formula(paste(neonics_to_eval[i],"~ per_corn * planting",
                            sep="")),data=osu_covariate)
  cat("\n",paste("Linear regression, ",neonics_to_eval[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
}
```
<br><br>

#####Not seeing any other significant relationships
<br><br><br>

Summary of initial findings
================================
<br>

* During corn planting periods only... there were greater concentrations of two neonic insecticides, **thiamethoxam** and **clothianidin**, and greater **total neonic concentrations** in field pollen when there was more corn planted nearby (within 2 km).
* Greater concentrations of these same neonics, as well as total neonic load, were associated with increased bee mortality.
* Presence of any detectable amount of these neonics, as well as any neonic in general, in pollen was associated with increased hive mortality.
* No significant relationships were found for other neonics measured in field pollen, or for any measurements from dead bees

<br><br>

